---
title: "Burlington Fire Department Incidents & Response Times, 2018 - 2019"
author: "Clay Crisholm, Ethan Lauricella & Lucía Carrera"
date: "10/17/2021"
output: html_document
---

## Data Background

```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Loading of packages we will use for our project
library(tidyverse)
library(lubridate)
library(leaflet)
library(gridExtra)

# Loading of the data frame fire to use
fire <- read.csv("O_fire_new.csv", na.strings = '', stringsAsFactors = TRUE)

# Result summaries of our data
tibble(fire)

# Visualization of the different variables available in our data set
names(fire)

## The names of our variables were summarized by r studio leading to names such as "Disp_d....m", to be able to clearly use our data we decided to modify these names for ease of use.

# Incident number
names(fire)[3] <- "Inc_num"

# Incident description
names(fire)[4] <- "Inc_des"

# Alarm date and time
names(fire)[5] <- "Alm_dt"

# Dispatch date and time
names(fire)[6] <- "Dis_dt"

# Dispatch Notified Date and Time
names(fire)[7] <- "Not_dt"

# Unit responding Date and Time
names(fire)[8] <- "Roll_dt"

# Arrival Date and Time
names(fire)[9] <- "Arv_dt"

# In Service Time
names(fire)[10] <- "Ins_dt"


##  We decided to add lattitude and longitude columns  
fire <- fire %>% mutate(geopoint2 = str_sub(geopoint, 1,-1)) %>%  
  separate(geopoint2, into = c('Lat', 'Long'), sep = ',')

fire <- fire %>% mutate(Long = as.numeric(Long), Lat =as.numeric(Lat))

names(fire)
```

```{r Time of Calls}
# The first graph we wanted to map out is the time of calls to see at what were the most "popular" times to call 
alarm <- fire %>% mutate(Call = mdy_hm(Alm_dt), 
                   Weekday = wday(Call, label = TRUE),
                   Mon = month(Call, label = TRUE),
                   Date = date(Call),
                   Hour = hour(Call))

# 
ggplot(data = alarm, 
       mapping = aes(x= Hour)) +
  geom_histogram(color = 'purple', fill = 'pink')
  

```


```{r Location of Calls}
#  Do a scatterplot of Latitude by Longitude, to see where calls are made in Burlington. Do you recognize the shape of the data?


# Location of Calls
ggplot(data = fire,
mapping = aes(x = Long, y = Lat )) +
  geom_point()

#  Using leaflet, make a map showing circles on the location of each incident.  Start here:
m <- leaflet() %>%
  addTiles() %>%
  setView(lat = 44.475, lng = -73.212, zoom = 12)

#  (Once you get the leaflet plot working, remove the scatterplot of Longitude by Latitude)
m  %>% addCircles(lng = fire$Long, lat = fire$Lat)

```

```{r Amount of times a Unit responds to call }
f2 <- fire %>% group_by(Unit) %>% summarize(count = n()) %>% arrange(desc(count)) %>% select(Unit = Unit, count = count)
f2
ggplot(data = fire, mapping = aes(x = Unit, fill = Unit) )+
  geom_bar()+ theme(axis.text.x = element_text(hjust=1, angle = 60)) + 
  scale_x_discrete(limits = f2$Unit)

```



```{r Type of incident description}
incident <- fire %>% group_by(Inc_des) %>% summarize(count = n()) %>% select(Inc_des = Inc_des, count = count) %>% arrange(desc(count)) %>% filter(count > 90)
incident
ggplot(data = fire, mapping = aes(x = Inc_des, fill = Inc_des))+
  geom_bar()+#theme(axis.text.x = element_text(hjust=1, angle = 60+  # We have an issue here
  scale_x_discrete(limits = incident$Inc_des)+theme(legend.position = 'none')

```
Around August the calls seem to double in frequency. This could be possible because the UVM student body is back in full swing.

```{r Month by shifts}
shift_time <- fire %>% filter(!is.na(Shift)) %>% mutate(Call = mdy_hm(Dis_dt),
                   Weekday = wday(Call, label = TRUE),
                   Mon = month(Call, label = FALSE),
                   Date = date(Call),
                   Hour = hour(Call))
names(shift_time)
ggplot(data = shift_time, mapping = aes(x= Mon, fill = Shift))+
  geom_histogram()


```

This plot shows the time spent at each incident in minutes. Most incidents only take about 20 minutes at the site. There is some cases where the shift is needed at a site for a long time. Most incidents are EMS calls, thus this makes sense as they most likely arrive at the scene and call for an ambulence to deliver the patient to a medical facility.

```{r Amount of time at each incident}

# Time spent at each site, A_ being the arrival time, and D_ being the departure time

inc_time <- fire %>% filter(!is.na(Alm_dt)) %>% filter(!is.na(Inc_des)) %>%
                    mutate(ACall = mdy_hm(Alm_dt), 
                   AWeekday = wday(ACall, label = TRUE),
                   AMon = month(ACall, label = TRUE),
                   ADate = date(ACall),
                   AHour = hour(ACall),
                   AMin = minute(ACall)) %>% 
                    mutate(DCall = mdy_hm(Ins_dt), 
                   DWeekday = wday(DCall, label = TRUE),
                   DMon = month(DCall, label = TRUE),
                   DDate = date(DCall),
                   DHour = hour(DCall),
                   DMin = minute(DCall))

inc_time <- inc_time %>% mutate(
                            time_workd = (DDate - ADate)*24*60,
                            time_workh = (DHour - AHour)*60,
                            time_workm = DMin - AMin,
                            total_m = time_workh + time_workm + time_workd)

inc_time <- inc_time %>% filter(!is.na(inc_time))


ggplot(data = inc_time, mapping = aes(x= total_m, fill = Shift))+
  geom_histogram(binwidth = 4) + labs(title = "Time spent at each incident site (minutes)")

```